---
title: "The Economic and Health Impacts of Severe Weather"
output: html_document
---

## Synopsis



## Data Processing

The raw data was downloaded from the URL provided on course website then loaded into R via read.csv function.  

```{r load data, cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "StormData.csv.bz2")
raw_data <- read.csv("StormData.csv.bz2")
```

Before 1996, NOAA only used this database to record tornado events. The purpose of this assignment is to compare human and economic impacts across event types, so I subset the analysis data to include only observations in the year 1996 and later (this first required converting date information from character class to POSIXct), and variables relevant to the assignment: event type ('EVTYPE'), fatalities, injuries, property damage ('PROPDMG' and 'PROPDMGEXP') and crop damage ('CROPDMG' and 'CROPDMGEXP'). Information regarding time, location, intensity, etc. are not relevant to the analysis of this project, so associated variables were removed from the analysis data. Additionally, I filtered out observations where casualties (fatalities and injuries) and damages (property or crop) were zero, since they would not contribute to the overall health or economic impact.  

```{r subset}
library(lubridate)
library(dplyr)
raw_data$BGN_DATE <- mdy_hms(raw_data$BGN_DATE)         #change date format to POSIXct

subset_data <- raw_data %>%                             #subset data
        filter(year(BGN_DATE) >= 1996) %>%              #choose observations from 1996 and later
        select(EVTYPE, FATALITIES:CROPDMGEXP) %>%       #choose relevant variables for analysis
        filter(FATALITIES != 0 | INJURIES !=0 | PROPDMG != 0 | CROPDMG != 0) #choose observations that resulted in casualty or damages
```

Next, I cleaned up the data. First, I converted the property and crop damage exponents into numbers that I could later manipulate arithmetically. 

```{r clean}
numeric_exp_data <- subset_data %>%                                     #convert exp symbols to numbers
        mutate(PROPDMGEXP = gsub("^", 0, PROPDMGEXP)) %>%
        mutate(PROPDMGEXP = gsub("K", 1000, PROPDMGEXP)) %>%
        mutate(PROPDMGEXP = gsub("M", 1000000, PROPDMGEXP)) %>%
        mutate(PROPDMGEXP = gsub("B", 1000000000, PROPDMGEXP)) %>%
        mutate(PROPDMGEXP = as.numeric(PROPDMGEXP)) %>%
        mutate(CROPDMGEXP = gsub("^", 0, CROPDMGEXP)) %>%
        mutate(CROPDMGEXP = gsub("K", 1000, CROPDMGEXP)) %>%
        mutate(CROPDMGEXP = gsub("M", 1000000, CROPDMGEXP)) %>%
        mutate(CROPDMGEXP = gsub("B", 1000000000, CROPDMGEXP)) %>%
        mutate(CROPDMGEXP = as.numeric(CROPDMGEXP))
```

I then multiplied the damage and damage exponent columns to create 'propertydamage' and 'cropdamage' variable that lists the damages as single numbers--much easier to work with. Finally, I created and selected the variable that are important for this analysis: casualties (fatalities plus injuries) which indicates harm to population's health, and economicimpact (property damage plus crop damage), while keeping EVTYPE. 

```{r cleaner}        
unified_numeric_data <- numeric_exp_data %>%
        mutate(propertydamage = PROPDMG * PROPDMGEXP) %>%
        mutate(cropdamage = CROPDMG * CROPDMGEXP) %>%
        mutate(economicimpact = propertydamage + cropdamage) %>%
        mutate(casualties = FATALITIES + INJURIES) %>%
        select(!(FATALITIES:cropdamage))
```



## Results
